Demographic consequences of partner diversity and turnover in a multi-species ant-plant mutualism

### 2025-06-25 Manuscript/ant_cactus_SUBMISSION2.tex

	\section*{Abstract} \label{res:abstract}
% 200 words max
The ecological consequences of sequential multi-species mutualisms, which involve multiple partners across time, with regards to partner diversity and turnover remain poorly understood.
In such sequential mutualisms, partner identity can change over time due to environmental variability or competitive displacement. 
This raises a key question: how does partner turnover influence host demography and long-term fitness?
We studied the fitness consequences of partner diversity and turnover in the sequential multi-species food-for-protection mutualism involving an EFN-bearing plant (tree cholla cactus \textit{Cylindropuntia imbricata}) and multiple species of nectar-feeding ants (\textit{Crematogaster opuntiae}, \textit{Liometopum apiculatum}, and other less frequent ants) that provide defense from herbivores. 
We used along-term dataset on plant vital rates and partner identity and Bayesian hierarchical statistical models to quantify the demographic consequences of association with different ant partners and the frequency and direction of turnover between mutually exclusive partner associations across the plant life cycle.
We then constructed a stochastic, multi-state Integral Projection Model (IPM) to simulate how different patterns of partner richness, composition, and sequential turnover affect population fitness under fluctuating environmental conditions.
We found that, while ant partners had different impacts on plant vital rates (\textit{C. opuntiae}-tended plants had advantages in growth and survival when small, and \textit{L. apiculatum}-tended plants had floral viability advantages), plant fitness was effectively insensitive to partner identity and number, because effects on the highest-sensitivity vital rates were consistent across partner species. 
Furthermore, there was no evidence that benefits of partner diversity manifest in the context of fluctuating environments (i.e., the portfolio effect). 
In contrast to much of the previous literature, this study highlights that demographic benefits of mutualism can be surprisingly robust to diversity and turnover of partners.

@@

%% Look at mechanisms which show how diversity impacts demography
In this study, we evaluate several mechanisms through which partner diversity might influence mutualistic function, drawing from broader biodiversity–ecosystem function theory \citep{Barrett2015,Ushio2020,Winfree2020}.
First, sampling effect occurs in systems with a distinct hierarchy of fitness effects (a consistent ranking of best to worst mutualists), when diversity increases the likelihood of including a high-quality partner \citep{Frederickson2013,Batstone2018}. \linelabel{res:mechanims_math} \linelabel{res:L45}
There is no benefit of diversity here \emph{per se}, rather, missed opportunity costs arise in systems with sequential partner associations, where less beneficial partners may preclude colonization by better mutualists \citep{Miller2007}.
Second, complementarity occurs in systems where partners differ functionally, when diversity increases the net benefits received by focal mutualist beyond what a single partner species could offer \citep{Batstone2018}.
Positive diversity effects through complementarity need not be additive: interference or synergies between partners can make their combined effect different than the expected from the sum of complementary functions \citep{Afkhami2014}. 
Even within a single mutualist guild, the benefits conferred by alternative partner species can vary by method , efficiency, or timing of benefit conferral \citep{Stachowicz2005,Bronstein2006,Stanton2003}. \linelabel{res:L53}
Third, portfolio effects stabilize host fitness across variable environments when different mutualists respond asynchronously to environmental fluctuations \citep{Batstone2018, Loreau1999, Lazaro2022,Ollerton2006,Alarcon2008}. \linelabel{res:portfolio_asynchronous} 
In this context, the portfolio effect implies that inter-annual variation in partner identity—especially if driven by environmental conditions—could stabilize overall plant fitness across years.
For example, if a high quality defensive partner is more likely to occur during dry years and a moderately defensive partner dominates in wet years, the shifting presence of these species could buffer the focal plant from herbivore damage under variable conditions. \linelabel{res:portfolio_theory}
Critically, portfolio effects require asynchronous environmental responses among partners, such that declines in one partner's benefits are offset by others, stabilizing the focal mutualist's fitness over time \citep{Loreau1999, Tilman1994, Hautier2014}.
This concept is well developed in biodiversity–ecosystem functioning literature, but has rarely been applied to mutualisms in which partner diversity is realized through turnover rather than coexistence.
While portfolio effects have been well studied in plant and animal communities, their role in mutualisms—particularly where partner diversity is sequential (turnover) rather than simultaneous—is less explored \citep{Ushio2020}.

%% Partner turnover
Partner diversity can affect a focal mutualist differently depending on whether partners are present simultaneously or sequentially (i.e., via partner turnover) \citep{Djieto-Lordon2005, Ness2006, Bruna2014, Barrett2015, Ushio2020, Dattilo2014}. 
In simultaneous associations partner interactions frequently lead to competition and interference \citep{Kiers2003,Batstone2018, Tgaard2015, Wulff2008}, reducing the net benefits received by the focal mutualist. \linelabel{res:L63}
Sequential associations have been shown to reduce these competitive partner-partner interactions, thereby boosting the focal mutualists benefits \citep{Wolfin2023,Stachowicz2005_2} and leading to more stable interactions \cite{Johnson2019}.
Turnover can happen at different timescales, from minutes to years \citep{Oliveira1999,Horvitz1986}. 
The timing, frequency, and directionality of partner turnover can influence the net fitness benefits to the host, particularly if turnover allows accumulation of complementary benefits or buffers environmental variability via portfolio effects \citep{Sachs2004,Fiala1994, Ushio2020}. 
If partner identity changes systematically across ontogeny, such as through succession of partners aligned with host life stages, this may further affect fitness outcomes \citep{Fonseca2003,Noe1994,Dejean2008}.
For example, plant susceptibility to enemies can change across life stages \citep{Boege2005,Barton2010}, so the benefits of a diverse guild of defensive mutualists are greatest when more defensive partner species align with more vulnerable life stages \citep{Djieto-Lordon2005,Dejean2008}.
Thus, the duration and order of interactions may be just as important as the identities of the interacting partners.

%% Defensive ant-plant mutualisms
Defensive ant-plant mutualisms -- where plants provide food and/or housing to ants that in turn defend them from enemies -- are widespread interactions that offer valuable model systems for the ecology and evolution of mutualism \citep{Bronstein1998, Bronstein2006}. 
Extrafloral nectar (EFN) -bearing plants can serve as dietary resources that promote ant abundance and colony size \citep{Byk2011, Ness2009, Ness2006, Donald2022}.
In turn, presence of defensive ant partners is often linked to reductions in herbivory  \citep{Trager2010, Rudgers2004} and demographic advantages for the plant partner \citep{Baez2016}.
Defensive ant-plant mutualisms are commonly multi-species, where a guild of ant partner species share, and often compete for, a plant mutualist \citep{Bronstein1998, Beattie1985, Trager2010, Agrawal1998}.
Ant partners can vary in their ability to deter herbivores \citep{Bruna2014}, and visitation by low quality ant partners can prevent visitation by higher quality partners, causing a reduction in fitness through missed opportunity costs \citep{Fraser2001, Frederickson2005}.
Susceptibility to herbivory can also vary significantly throughout the life stages of the plant \citep{Boege2005}, suggesting that the order and timing of successive partners is important to the fitness impacts of the combined partner guild \citep{Barton2010, Boege2005, Fonseca2003}.
Finally, herbivore identity and pressure can vary inter-annually \citep{Wetzel2023}, much like mutualist identity and presence, meaning the threat plants face can vary just as much as the protection they receive due to temporal stochasticity. 
Previous studies have investigated how ant partner diversity affects plant fitness \citep{Palmer2010,Fiala1994,Gaume1998,Dattilo2014,Ludka2015}. 
However, little is known about the combined effects of partner identity, directional partner turnover, and temporal stochasticity, particularly because the long-term data needed to quantify fluctuations in ant-plant interactions are rarely available. 
Specifically, sequential associations require much longer-term data in order to analyze due to an accumulation of partners across time rather than the ability to capture multiple at once. \linelabel{res:long-term-data}

%% What this study set out to do
This study examined the consequences of partner diversity in a food-for-protection mutualism between the tree cholla cactus (\textit{Cylindropuntia imbricata}), a long-lived EFN-bearing plant, and multiple species of ant partners.
Tree cholla are tended by two common ant partners and several additional rarer species, all of which collect EFN in exchange for defense against herbivores.
These ant species locally co-occur but individual plants are typically tended by only one species that patrols the plant around-the-clock and maintains control of the plant's nectar resources, usually for an entire growing season \citep{Ohm2014, Donald2022}. 
Switches between partner species, or between vacancy and ant occupancy, commonly occur from one growing season to the next \citep{Miller2007}. 
Prior experiments suggested a hierarchy of mutualist quality, with \textit{Liometopum apiculatum} providing strong anti-herbivore defense and \textit{Crematogaster opuntiae} having net negative effects because herbivore deterrence is outweighed by deterrence of pollinators \citep{Miller2007,Ohm2014}.  \linelabel{res:L111}
Even the basic quantification of how different ant species affect host demography across the plant life cycle is rarely possible, particularly in long-lived perennials. 
Most studies evaluate short-term fitness proxies like seed set or herbivory at a single life stage \citep{Palmer2010, Bruna2014, Trager2010}, while few trace demographic effects across multiple stages.
Similarly, while turnover of mutualist partners is known to occur \citep{Ness2006, Djieto-Lordon2005}, long-term data documenting patterns and consequences of partner switching are uncommon, especially in field systems. \linelabel{res:long-term-data3}
Thus, this study provides unique empirical insight into how mutualist quality and temporal partner dynamics shape host population trajectories.

%% Models and questions
Here we used a unique long-term data set that allows us to explore mutualistic associations with multiple partner species, longitudinal dynamics of partner turnover, and demographic effects of alternative partners that vary across plant size structure and across nearly 20 years of inter-annual fluctuations. 
The long-term nature of this observational dataset gives us the power to investigate whether and through which mechanism(s) partner diversity affects the fitness benefits of ant visitation. \linelabel{res:long-term-data2}
Specifically, we asked:
	\begin{enumerate}	
		\item{What are the vital rate effects of association with alternative partners and how do these effects fluctuate across years?}
		\item{What are the frequency and direction of partner turnover across the plant life cycle?}	
		\item{What is the net effect of partner diversity on plant fitness, and what mechanism(s) explain(s) this effect?}
	\end{enumerate}
To answer these questions we used a hierarchical Bayesian statistical approach to estimate demographic vital rates for hosts in different states of ant occupancy and to quantify state-dependent partner turnover from the long-term data. 
We then used a stochastic, multi-state integral projection model (IPM) that combines diverse effects on vital rates and pathways of partner turnover to quantify effects of partner diversity on plant fitness.

>>>>

	\section*{Abstract} \label{res:abstract}
% 200 words max
The ecological consequences of sequential multi-species mutualisms, which involve multiple partners across time, with regards to partner diversity and turnover remain poorly understood.
In such sequential mutualisms, partner identity can change over time due to environmental variability or competitive displacement. 
This raises a key question: how does partner turnover influence host demography and long-term fitness?
We studied the fitness consequences of partner diversity and turnover in the sequential multi-species food-for-protection mutualism involving an EFN-bearing plant (tree cholla cactus \textit{Cylindropuntia imbricata}) and multiple species of nectar-feeding ants (\textit{Crematogaster opuntiae}, \textit{Liometopum apiculatum}, and other less frequent ants) that provide defense from herbivores. 
We used a long-term dataset on plant vital rates and partner identity and Bayesian hierarchical statistical models to quantify the demographic consequences of association with different ant partners and the frequency and direction of turnover between mutually exclusive partner associations across the plant life cycle.
We then constructed a stochastic, multi-state Integral Projection Model (IPM) to simulate how different patterns of partner richness, composition, and sequential turnover affect population fitness under fluctuating environmental conditions.
We found that, while ant partners had different impacts on plant vital rates (\textit{C. opuntiae}-tended plants had advantages in growth and survival when small, and \textit{L. apiculatum}-tended plants had floral viability advantages), plant fitness was effectively insensitive to partner identity and number, because effects on the highest-sensitivity vital rates were consistent across partner species. 
Furthermore, there was no evidence that benefits of partner diversity manifest in the context of fluctuating environments (i.e., the portfolio effect). 
In contrast to much of the previous literature, this study highlights that demographic benefits of mutualism can be surprisingly robust to diversity and turnover of partners.

@@

%% Look at mechanisms which show how diversity impacts demography
In this study, we evaluate several mechanisms through which partner diversity might influence mutualistic function, drawing from broader biodiversity–ecosystem function theory \citep{Barrett2015,Ushio2020,Winfree2020}.
First, sampling effect occurs in systems with a distinct hierarchy of fitness effects (a consistent ranking of best to worst mutualists), when diversity increases the likelihood of including a high-quality partner \citep{Frederickson2013,Batstone2018}. \linelabel{res:mechanims_math} \linelabel{res:L45}
There is no benefit of diversity here \emph{per se}, rather, missed opportunity costs arise in systems with sequential partner associations where less beneficial partners may preclude colonization by better mutualists \citep{Miller2007}.
Second, complementarity occurs in systems where partners differ functionally, when diversity increases the net benefits received by focal mutualist beyond what a single partner species could offer \citep{Batstone2018}.
Positive diversity effects through complementarity need not be additive: interference or synergies between partners can make their combined effect different than expected from the sum of complementary functions \citep{Afkhami2014}. 
Even within a single mutualist guild, the benefits conferred by alternative partner species can vary by method, efficiency, or timing of benefit conferral \citep{Stachowicz2005,Bronstein2006,Stanton2003}. \linelabel{res:L53}
Third, portfolio effects stabilize host fitness across variable environments when different mutualists respond asynchronously to environmental fluctuations \citep{Batstone2018, Loreau1999, Lazaro2022,Ollerton2006,Alarcon2008}. \linelabel{res:portfolio_asynchronous} 
In this context, the portfolio effect implies that inter-annual variation in partner identity—especially if driven by environmental conditions—could stabilize overall plant fitness across years.
For example, if a high quality defensive partner is more likely to occur during dry years and a moderately defensive partner dominates in wet years, the shifting presence of these species could buffer the focal plant from herbivore damage under variable conditions. \linelabel{res:portfolio_theory}
Critically, portfolio effects require asynchronous environmental responses among partners, such that declines in one partner's benefits are offset by others, stabilizing the focal mutualist's fitness over time \citep{Loreau1999, Tilman1994, Hautier2014}.
This concept is well developed in biodiversity–ecosystem functioning literature, but has rarely been applied to mutualisms in which partner diversity is realized through turnover rather than coexistence.
While portfolio effects have been well studied in plant and animal communities, their role in mutualisms—particularly where partner diversity is sequential (turnover) rather than simultaneous—is less explored \citep{Ushio2020}.

%% Partner turnover
Partner diversity can affect a focal mutualist differently depending on whether partners are present simultaneously or sequentially (i.e., via partner turnover) \citep{Djieto-Lordon2005, Ness2006, Bruna2014, Barrett2015, Ushio2020, Dattilo2014}. 
In simultaneous associations, partner interactions frequently lead to competition and interference \citep{Kiers2003,Batstone2018, Tgaard2015, Wulff2008}, reducing the net benefits received by the focal mutualist. \linelabel{res:L63}
Sequential associations have been shown to reduce these competitive partner-partner interactions, thereby boosting the focal mutualists benefits \citep{Wolfin2023,Stachowicz2005_2} and leading to more stable interactions \cite{Johnson2019}.
Turnover can happen at different timescales, from minutes to years \citep{Oliveira1999,Horvitz1986}. 
The timing, frequency, and directionality of partner turnover can influence the net fitness benefits to the host, particularly if turnover allows accumulation of complementary benefits or buffers environmental variability via portfolio effects \citep{Sachs2004,Fiala1994, Ushio2020}. 
If partner identity changes systematically across ontogeny, such as through succession of partners aligned with host life stages, this may further affect fitness outcomes \citep{Fonseca2003,Noe1994,Dejean2008}.
For example, plant susceptibility to enemies can change across life stages \citep{Boege2005,Barton2010}, so the benefits of a diverse guild of defensive mutualists are greatest when more defensive partner species align with more vulnerable life stages \citep{Djieto-Lordon2005,Dejean2008}.
Thus, the duration and order of interactions may be just as important as the identities of the interacting partners.

%% Defensive ant-plant mutualisms
Defensive ant-plant mutualisms -- where plants provide food and/or housing to ants that in turn defend them from enemies -- are widespread interactions that offer valuable model systems for the ecology and evolution of mutualism \citep{Bronstein1998, Bronstein2006}. 
Extrafloral nectar (EFN) -bearing plants can serve as dietary resources that promote ant abundance and colony size \citep{Byk2011, Ness2009, Ness2006, Donald2022}.
In turn, presence of defensive ant partners is often linked to reductions in herbivory  \citep{Trager2010, Rudgers2004} and demographic advantages for the plant partner \citep{Baez2016}.
Defensive ant-plant mutualisms are commonly multi-species, where a guild of ant partner species share, and often compete for, a plant mutualist \citep{Bronstein1998, Beattie1985, Trager2010, Agrawal1998}.
Ant partners can vary in their ability to deter herbivores \citep{Bruna2014} and visitation by low quality ant partners can prevent visitation by higher quality partners, causing a reduction in fitness through missed opportunity costs \citep{Fraser2001, Frederickson2005}.
Susceptibility to herbivory can also vary significantly throughout the life stages of the plant \citep{Boege2005}, suggesting that the order and timing of successive partners is important to the fitness impacts of the combined partner guild \citep{Barton2010, Boege2005, Fonseca2003}.
Finally, herbivore identity and pressure can vary inter-annually \citep{Wetzel2023}, much like mutualist identity and presence, meaning the threat plants face can vary just as much as the protection they receive due to temporal stochasticity. 
Previous studies have investigated how ant partner diversity affects plant fitness \citep{Palmer2010,Fiala1994,Gaume1998,Dattilo2014,Ludka2015}, however little is known about the combined effects of partner identity, directional partner turnover and temporal stochasticity, particularly because the long-term data needed to quantify fluctuations in ant-plant interactions are rarely available. 
Specifically, sequential associations require much longer-term data in order to analyze, due to an accumulation of partners across time rather than the ability to capture multiple at once. \linelabel{res:long-term-data}

%% What this study set out to do
This study examined the consequences of partner diversity in a food-for-protection mutualism between the tree cholla cactus (\textit{Cylindropuntia imbricata}), a long-lived EFN-bearing plant, and multiple species of ant partners.
Tree cholla are tended by two common ant partners and several additional rarer species, all of which collect EFN in exchange for defense against herbivores.
These ant species locally co-occur, but individual plants are typically tended by only one species that patrols the plant around-the-clock and maintains control of the plant's nectar resources, usually for an entire growing season \citep{Ohm2014, Donald2022}. 
Switches between partner species, or between vacancy and ant occupancy, commonly occur from one growing season to the next \citep{Miller2007}. 
Prior experiments suggested a hierarchy of mutualist quality, with \textit{Liometopum apiculatum} providing strong anti-herbivore defense and \textit{Crematogaster opuntiae} having net negative effects because herbivore deterrence is outweighed by deterrence of pollinators \citep{Miller2007,Ohm2014}.  \linelabel{res:L111}
Even the basic quantification of how different ant species affect host demography across the plant life cycle is rarely possible, particularly in long-lived perennials. 
Most studies evaluate short-term fitness proxies like seed set or herbivory at a single life stage \citep{Palmer2010, Bruna2014, Trager2010}, while few trace demographic effects across multiple stages.
Similarly, while turnover of mutualist partners is known to occur \citep{Ness2006, Djieto-Lordon2005}, long-term data documenting patterns and consequences of partner switching are uncommon, especially in field systems. \linelabel{res:long-term-data3}
Thus, this study provides unique empirical insight into how mutualist quality and temporal partner dynamics shape host population trajectories.

%% Models and questions
Here we used a unique long-term data set that allows us to explore mutualistic associations with multiple partner species, longitudinal dynamics of partner turnover and demographic effects of alternative partners that vary across plant size structure and across nearly 20 years of inter-annual fluctuations. 
The long-term nature of this observational dataset gives us the power to investigate whether, and through which mechanism(s), partner diversity affects the fitness benefits of ant visitation. \linelabel{res:long-term-data2}
Specifically, we asked:
	\begin{enumerate}	
		\item{What are the vital rate effects of association with alternative partners and how do these effects fluctuate across years?}
		\item{What are the frequency and direction of partner turnover across the plant life cycle?}	
		\item{What is the net effect of partner diversity on plant fitness, and what mechanism(s) explain(s) this effect?}
	\end{enumerate}
To answer these questions we used a hierarchical Bayesian statistical approach to estimate demographic vital rates for hosts in different states of ant occupancy and to quantify state-dependent partner turnover from the long-term data. 
We then used a stochastic, multi-state integral projection model (IPM) which combines diverse effects on vital rates and pathways of partner turnover to quantify effects of partner diversity on plant fitness.
